Supplementary materials for: Modelling palaeoecological time series using generalized additive models

### 2018-08-15 supplementary-materials.Rmd
revise the materials

---
title: "Supplementary materials for: Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: May 13, 2018
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    #template: my.latex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

@@

The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, and *tidyr*. Also, the *schoenberg* package is required; it is not on CRAN but can be installed from Github.

```{r load-packages, cache = FALSE}
library("mgcv")
library("scam")
library("ggplot2")
library("cowplot")

## schoenberg is not on CRAN, install from github:
## install.packages("devtools")
## devtools::install_github("gavinsimpson/schoenberg")
library("schoenberg")
library("tidyr")

## Default ggplot theme
theme_set(theme_bw())
```

The example data sets are also stored on github; https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling. Once downloaded the data are read in and processed a little

@@

## load braya so data set
braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
                    skip = 84)
names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung",
                  "YearOld", "UK37")
braya <- transform(braya, sampleInterval = YearYoung - YearOld)
head(braya)

@@

@@

```{r fit-small-water-gamm}
## fit small water GAM usin gamm() with a CAR(1)
mod <- gamm(d15N ~ s(Year, k = 15), data = small,
            correlation = corCAR1(form = ~ Year), method = "REML")
```

@@

The model summary is prepared from the `$gam` component of the fitted model

@@

The output shows the estimated complexity of the fitted smooth, in terms of the effective degrees of freedom of the spline. An associated $F$ statistic and test of the null hypothesis of no trend (effect). Here the estimated trend provides strong evidence against this null.

@@

```{r car1-plot}
## plot CAR(1) process
maxS <- with(small, diff(range(Year))) ## too large, truncate to 50
S <- seq(0, 50, length = 100)

car1 <- setNames(as.data.frame(t(outer(smallPhi, S, FUN = `^`)[1, , ])),
                 c("Lower","Correlation","Upper"))
car1 <- transform(car1, S = S)

@@

The same model is fitted to the Braya-Sø data set. Note however that in order to even fit the model with both a smooth and the CAR(1) process, I have had to change the default optimiser used to fit the model, and reduce the basis dimension to a small number. We also fit the model using GCV, which is the defaut, hence no `method` argument

@@

The next couple of code chunks prepare plots of the fitted GAMS. The general idea is to predict fom the fitted model for a fine grid of points over the range of the time variable. First we plot the trend for Small Water with an approximate 95% confidence interval assuming asymptotic normality

```{r small-plot-fitted-models}
N <- 300   # number of points at which to evaluate the splines
## Predict from the fitted model
newYear <- with(small, data.frame(Year = seq(min(Year), max(Year),
                                             length.out = 200)))
newYear <- cbind(newYear,
                 data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
newYear <- transform(newYear,
                     upper = fit + (2 * se.fit),
                     lower = fit - (2 * se.fit))

## Plot simulated trends
small_fitted <- ggplot(newYear, aes(x = Year, y = fit)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, x = Year), alpha = 0.2,
                inherit.aes = FALSE, fill = "black") +
    geom_point(data = small, mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE) +
    geom_line() +
    labs(y = d15n_label, x = "Year CE")
small_fitted
```

@@

```{r braya-plot-fitted-models}
## ggplot with data and fitted spline, then resids vs time in second panel
newBraya <- with(braya, data.frame(Year = seq(min(Year), max(Year),
                                              length.out = N)))
newBraya <- cbind(newBraya,
                  data.frame(predict(braya.car1$gam, newBraya,
                                     se.fit = TRUE)))
crit.t <- qt(0.975, df = df.residual(braya.car1$gam))
newBraya <- transform(newBraya,
                      upper = fit + (crit.t * se.fit),
                      lower = fit - (crit.t * se.fit))
## add GAM GCV results
fit_gcv <- predict(braya.gcv, newdata = newBraya, se.fit = TRUE)
newBraya <- rbind(newBraya, newBraya) # extend newBraya to take GCV results
newBraya[seq(N+1, length.out = N, by = 1), ]$fit <- fit_gcv$fit
newBraya[seq(N+1, length.out = N, by = 1), ]$upper <-
    fit_gcv$fit + (qt(0.975, df.residual(braya.gcv)) * fit_gcv$se.fit)
newBraya[seq(N+1, length.out = N, by = 1), ]$lower <-
    fit_gcv$fit - (qt(0.975, df.residual(braya.gcv)) * fit_gcv$se.fit)
newBraya <- transform(newBraya,
                      Method = rep(c("GAMM (CAR(1))", "GAM (GCV)"), 
                                   each = N))

@@

```{r manuscript-fig-2}
plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

To proceed with the Braya-Sø example, we need to increase the basis dimension (`k = 45`), fit using `method = "REML"`, and use observational weights. Here I use the `sampleInterval` variable as the measure of lake years per sample, and to avoid changing the model likelihood, the weights are actually the values of `sampleInterval` divided by the mean of `sampleInterval`:

```{r final-braya-reml-fit}
## TPRS, weights as sampleInterval
braya_reml <- gam(UK37 ~ s(Year, k = 45, bs = "tp"), data = braya,
                  method = "REML",
                  weights = sampleInterval / mean(sampleInterval))
```

@@

```{r figure-7}
plot_grid(smallSim.plt, brayaSim.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

Across-the-function and simultaneous confidence intervals are computed using the `confint()` method. The type of interval required is given via the `type` argument with options `"confidence"` and `"simultaneous"`.

```{r small-compare-intervals}
## small water
sw.cint <- confint(mod, parm = "Year", newdata = newYear,
                   type = "confidence")
sw.sint <- confint(mod, parm = "Year", newdata = newYear,
                   type = "simultaneous")

smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = est)) +
    geom_ribbon(data = sw.sint,
                mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = d15n_label, x = "Year CE")
smallInt.plt
```

```{r braya-compare-intervals}
## braya so
bs.cint <- confint(braya_reml, parm = "Year", newdata = newBraya,
                   type = "confidence")
bs.sint <- confint(braya_reml, parm = "Year", newdata = newBraya,
                   type = "simultaneous")

brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = est)) +
    geom_ribbon(data = bs.sint,
                mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = braya_ylabel, x = "Year CE")
brayaInt.plt
```

Figure 8 in the manuscript was prepared using

```{r figure-8}
plot_grid(smallInt.plt, brayaInt.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

```{r profile-range-parameter-in-gp-smooth-model}
nn <- 200    # number of points at which to evaluate profile likelihood
dseq <- seq(15, 500, length.out = nn)  # effective ranges to fit at
Mat <- SEx <- numeric(length = nn)     # object to hold model fits
for (i in seq_along(dseq)) { 
    ## iterate over dseq, fit GP GAM w Matérn covariance
    Mat[i] <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, dseq[i])),
                  weights = sampleInterval / mean(sampleInterval),
                  data = braya, method = "REML",
                  family = gaussian())[["gcv.ubre"]]
    ## fit squared exponential
    SEx[i] <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, dseq[i], 1)),
                  weights = sampleInterval / mean(sampleInterval),
                  data = braya, method = "REML",
                  family = gaussian())[["gcv.ubre"]]
}

@@

```{r fit-example-gp-smooths, dependsons = "profile-range-parameter-in-gp-smooth-model"}
effRange1 <- 250      # sets effective range in years for Matérn correl
## minima from profile likelihood
effRange2 <- with(subset(reml.scr, cor == "Matérn"), dseq[which.min(reml)])
effRange3 <- with(subset(reml.scr, cor == "Exponential"), dseq[which.min(reml)])

## Matern
gp2 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, effRange2)),
           data = braya,
           method = "REML", weights = sampleInterval / mean(sampleInterval))
## Power exponential
gp3 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, effRange3, 1)),
           data = braya,
           method = "REML", weights = sampleInterval / mean(sampleInterval))

newd <- with(braya, data.frame(Year = seq(min(Year), max(Year),
                               length.out = 1000)))
p.gp2 <- transform(newd,
                   fitted = predict(gp2, newdata = newd, type = "response"),
                   effRange = round(effRange2))
p.gp3 <- transform(newd,
                   fitted = predict(gp3, newdata = newd, type = "response"),
                   effRange = round(effRange3))
## pred <- rbind(p.gp1, p.gp2, p.gp3)
pred <- rbind(p.gp2, p.gp3)
pred <- transform(pred, effRange = factor(effRange),
                  cor = rep(c("Matérn", "Exponential"), each = nrow(newd)))
```

The estimated trends are plotted using

```{r plot-both-gp-smooth-trends}
## plot at two values of h
gp.plt2 <- ggplot(pred, aes(x = Year, y = fitted, colour = cor)) +
    geom_line() + theme(legend.position = "right") +
    geom_point(aes(x = Year, y = UK37), data = braya, inherit.aes = FALSE) +
    scale_colour_manual(name = "", values = c("#e66101","#5e3c99")) +
    labs(y = braya_ylabel, x = "Year CE")
```

@@

```{r figure-12}
plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"),
          align = "hv", axis = "lr")
```

@@

```{r adaptove-smooth}
## Adaptive spline, weights as sampleInterval
mod_ad <- gam(UK37 ~ s(Year, k = 45, bs = "ad"), data = braya,
              method = "REML",
              weights = sampleInterval / mean(sampleInterval))
```

# Compare various trends

@@

1. Thin plate regression spline,
2. Gaussian process spline (Matérn correlation functions), and
3. Adaptive smoother

@@

```{r braya-so-model-comparisons, dependson = "fit-example-gp-smooths"}
## model it using gam()
effRange <- effRange2

## TPRS, weights as sampleInterval, k = needs to be higher
mod_tprs <- gam(UK37 ~ s(Year, k = 45, bs = "tp"), data = braya,
                method = "REML",
                weights = sampleInterval / mean(sampleInterval))

@@

```{r, dependson = -1}
## wrapper to simulate from a fitted GAM with the arguments/settings
## I want
simulateSWModels <- function(mod, newdata, nsim, seed = 42) {
    sims <- simulate(mod, nsim = nsim, newdata = newdata, seed = seed)
    as.vector(sims)
}

@@

plot_grid(plt1, plt2, plt3, ncol = 1, labels = "auto", align = "hv",
          axis = "lrtb", rel_widths = c(0.5, 1, 1))
```

>>>>

---
title: "Supplementary materials for: Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: August 14, 2018
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    highlight: tango
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

@@

The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, and *tidyr*.

```{r load-packages, cache = FALSE}
library("mgcv")
library("scam")
library("ggplot2")
library("cowplot")
library("tidyr")
```
In addition, the *schoenberg* package is required; this in-development package is not on CRAN but can be installed directly from GitHub using functions from the *devtools* package. To install the package, install *devtools* and then use `devtools::install_github()` to install *schoenberg*, as shown below:

```{r install-schoenberg, eval = FALSE}
## schoenberg is not on CRAN, install from github:
install.packages("devtools")
devtools::install_github("gavinsimpson/schoenberg")
```
Assuming that the installation of *schoenberg* completes without error, the package can be loaded as usual
```{r load-schoenberg, cache = FALSE}
library("schoenberg") # need to change the name of the package
```

The final environment-preparation step is to set the default *ggplot* theme, which the loading of *cowplot* has over-ridden. Here I use the more-minimal black-and-white theme (`theme_bw()`)

```{r ggplot-default-theme, cache = FALSE}
## Default ggplot theme
theme_set(theme_bw())
```

# Load the data sets

The example data sets are also stored on GitHub; https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling. Once downloaded the data are read in and processed a little

@@

## load braya so data set
braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
                    skip = 84)
## clean up variable names
names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung",
                  "YearOld", "UK37")
## add a variable for the amount of time per sediment sample
braya <- transform(braya, sampleInterval = YearYoung - YearOld)
head(braya)

@@

Instead of following the structure of the paper exactly, the subsequent sections focus on the two example data sets in turn, beginning with Small Water.

## Small Water

A GAM is typically fitted using the `gam()` function from the *mgcv* package. The typical call includes a formula describing the response variable and the linear predictor separated by the `~` (tilde) symbol. The linear predictor part of the model contains the smooth function of the time variable in the data set, and is indicated using the `s()` function. Unless other options are needed, we only need specify where the variables in the formula can be found via the `data` argument, and that we wish to use REML smoothness selection.

Putting this together we have the following function call to fit a simple GAM to the Small Water isotope data

```{r fit-small-water-gam}
m <- gam(d15N ~ s(Year, k = 15), data = small, method = "REML")
```

As discussed in the paper, this model assumes that residuals[^1] are independent. To account for residual temporal autocorrelation we can include a continuous-time first-order autoregressive (CAR(1)) process in the model residuals for GAMs fitted to data that are conditionally distributed Gaussian.

[^1]: or equivalently that the observations, conditional upon the model, are independent.

@@

```{r fit-small-water-gamm}
## fit small water GAM using gamm() with a CAR(1)
mod <- gamm(d15N ~ s(Year, k = 15), data = small,
            correlation = corCAR1(form = ~ Year), method = "REML")
```

@@

The object returned by `gamm()` includes both the linear mixed model and GAM faces of the model, and as a result we need to access the separate elements (`lme` and `gam` respectively) when proceeding to explore the model fit. The model summary is prepared from the `$gam` component of the fitted model

@@

The output shows the estimated complexity of the fitted smooth, expressed in terms of the effective degrees of freedom of the spline. An associated $F$ statistic and test of the null hypothesis of no trend (effect) are also shown. Here the estimated trend provides strong evidence against this null.

@@

```{r car1-plot, dependson = 'extract-phi'}
## plot CAR(1) process
S <- seq(0, 50, length = 100)
car1 <- setNames(as.data.frame(t(outer(smallPhi, S, FUN = `^`)[1, , ])),
                 c("Lower","Correlation","Upper"))
car1 <- transform(car1, S = S)

@@

The next code chunk prepares a plot of the fitted GAMS. The general idea is to predict from the fitted model for a fine grid of points over the range of the time variable. Below I plot the trend for Small Water with an approximate 95% point-wise confidence interval that assumes asymptotic normality

```{r small-plot-fitted-models}
N <- 300   # number of points at which to evaluate the smooth

## create new data to predict at; 200 evenly-spaced values over `Year`
newYear <- with(small, data.frame(Year = seq(min(Year), max(Year),
                                             length.out = 200)))
## Predict from the fitted model; note we predict from the $gam part
newYear <- cbind(newYear,
                 data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
## Create the confidence interval
crit.t <- qt(0.975, df = df.residual(mod$gam))
newYear <- transform(newYear,
                     upper = fit + (crit.t * se.fit),
                     lower = fit - (crit.t * se.fit))

## Plot estimated trend
small_fitted <- ggplot(newYear, aes(x = Year, y = fit)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, x = Year), alpha = 0.2,
                inherit.aes = FALSE, fill = "black") +
    geom_point(data = small, mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE) +
    geom_line() +
    labs(y = d15n_label, x = "Year CE")
small_fitted
```

## Braya-Sø

The same model is fitted to the Braya-Sø data set. Note however that in order to even fit the model with both a smooth and the CAR(1) process, I have had to change the default optimiser used to estimate the model parameters, and reduce the basis dimension to a small number. We also fit the model using GCV, which is the defaut, hence no `method` argument

@@

@@

```{r braya-plot-fitted-models}
N <- 300   # number of points at which to evaluate the smooth

## data to predict at
newBraya <- with(braya, data.frame(Year = seq(min(Year), max(Year),
                                              length.out = N)))
## add predictions from GAMM + CAR(1) model
newBraya <- cbind(newBraya,
                  data.frame(predict(braya.car1$gam, newBraya,
                                     se.fit = TRUE)))
crit.t <- qt(0.975, df = df.residual(braya.car1$gam))
newBraya <- transform(newBraya,
                      upper = fit + (crit.t * se.fit),
                      lower = fit - (crit.t * se.fit))

## add GAM GCV results
fit_gcv <- predict(braya.gcv, newdata = newBraya, se.fit = TRUE)
crit.t <- qt(0.975, df.residual(braya.gcv))
newGCV <- data.frame(Year   = newBraya[["Year"]],
                     fit    = fit_gcv$fit,
                     se.fit = fit_gcv$se.fit)
newGCV <- transform(newGCV,
                    upper = fit + (crit.t * se.fit),
                    lower = fit - (crit.t * se.fit))
newBraya <- rbind(newBraya, newGCV)                # bind on GCV results
## Add indicator variable for model
newBraya <- transform(newBraya,
                      Method = rep(c("GAMM (CAR(1))", "GAM (GCV)"), 
                                   each = N))

@@

```{r manuscript-fig-2, fig.show = "hide"}
plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

### Checking if the size of the basis expansion is sufficient

Clearly we are not able to identify both the wiggly trend and the CAR(1) process in a single model. We might proceed by fitting a simple GAM via `gam()` with REML smoothness selection and moderate number of basis functions --- here we set `k = 15` for illustration.

```{r braya-fit-k-too-small}
braya_low_k <- gam(UK37 ~ s(Year, k = 15), data = braya, method = "REML")
```

Before proceeding, we should perform a check to determine if the number of basis functions requested is sufficient to capture the wiggliness in the data. This test is provided by `gam.check()`

```{r check-braya-fit-k-too-small, dependson = -1, fig.show = "hide"}
gam.check(braya_low_k)
```

In the output, the column labelled `k'` is the size of the basis used (the number of basis functions) --- note that it is 14 and not the requested 15 functions as the constant basis function has been removed as it is confounded with the model constant term, the intercept. The column labelled `edf` indicates the *effective degrees of freedom* for the estimated smooth. The `k-index` column contains the value of a test statistic for a test of sufficient number of basis functions --- ideally, we'd like the value of `k-index` to be close to or greater than 1. The quoted *p* value measures the support for the null hypothesis that enough basis functions were available. In practial terms, the low `k-index` and very low *p* value suggest that the requested number of basis functions used to fit the model was too low.

The solution is to increase `k`, say by doubling the original value, and refit the model and perform the basis dimension check using `gam.check()`, repeating this process until `k-index` is close to or greater than 1 and the *p* value is larger than a desired threshold. In my experience, it may not be possible with some data sets to increase `k` large enough to result in a `k-index` > 1 and a *p* value > 0.05, given the available. It is important to remember that the test provided by `gam.check()` is only an heuristic one and shouldn't be seen as infallible. If you find yourself increasing `k` to large values relative to the number of samples in your data set without achieving a `k-index` > 1 or *p* > 0.05, then note the value in the `edf` column as you increase `k`. If the `edf` of the model changes little as you continue to increase `k`, this may be evidence that the test is failing in this particular instance and that the estimated smooth is not dependent on the value of `k` chosen.

## Accounting for heteroscedasticity due to time averaging

To proceed with the Braya-Sø example, we need to increase the basis dimension (`k = 45`), fit using `method = "REML"`. As discussed in the manuscript, we should also account for the non-constant variance, or *heteroscedasticity*, that may arise due to variation in the amount of time (or "lake years") that each sample represents. All else equal, we expect that samples that average more time have smaller variance than those that average a smaller amount of time. To include information about the expected heteroscedasticity in the GAM we can use observational weights[^2]. Here I use the `sampleInterval` variable that I created earlier as the measure of lake years per sample, and to avoid changing the model likelihood, the weights use are the values of `sampleInterval` divided by the mean of `sampleInterval`:

```{r final-braya-reml-fit}
braya_reml <- gam(UK37 ~ s(Year, k = 45), data = braya,
                  method = "REML",
                  weights = sampleInterval / mean(sampleInterval))
summary(braya_reml)
```

[^2]: This simple solution is for a Gaussian GAM. For models with other distributions, the exact interpretation of observational weights will vary. For such data, where the conditional mean and variance of the observations are linked, the use of location-scale, or distributional, models may be more apporpriate and informative.

If we now check if 44 (`k = 45`) basis functions is sufficient
```{r final-braya-gam-check, fig.show = "hide"}
gam.check(braya_reml)
```
we note that the value of `k-index` is now greater than 1 and the *p* value suggests there is very little evidence against the null hypothesis. Also note that the `edf` of the estimated smooth is considerably below the maximum (`k'`). This is the effect of the smoothness penalty removing the excessive wiggliness possible with 44 basis functions. If the `edf` were close to `k'`, you might consider increasing `k` by a modest amount (in this instance by 5 or 10 additional basis functions) to assure yourself that you have sufficient basis functions. This has the additional advantage of allowing a richer set of smooths of a given complexity (`edf`) to be represented using the basis functions, which may improve the model fit to the data without changing the `edf` too much.

@@

```{r figure-7, fig.show = "hide"}
plot_grid(smallSim.plt, brayaSim.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

Across-the-function and simultaneous confidence intervals are computed using the `confint()` method. The type of interval required is given via the `type` argument with options `"confidence"` and `"simultaneous"`. Fro example, for Small Water we would use

```{r small-compare-intervals}
sw.cint <- confint(mod, parm = "Year", newdata = newYear,
                   type = "confidence")
sw.sint <- confint(mod, parm = "Year", newdata = newYear,
                   type = "simultaneous")
head(sw.sint)
```

The `confint()` methods return data frames suitable for subsequent plotting with **ggplot**. The columns labelled `est` and `se` are the estimate values of the smooth and its standard error, respectively. The variables `lower` and `upper` contain the values of the lower and upper bounds on the requested interval. The intervals can be plotted as follows

```{r plot-small-water-intervals}
smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = est)) +
    geom_ribbon(data = sw.sint,
                mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = d15n_label, x = "Year CE")
smallInt.plt
```

The intervals for Braya-Sø are created
```{r braya-compare-intervals}
bs.cint <- confint(braya_reml, parm = "Year", newdata = newBraya,
                   type = "confidence")
bs.sint <- confint(braya_reml, parm = "Year", newdata = newBraya,
                   type = "simultaneous")
```
and plotted
```{r plot-braya-so-intervals}
brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = est)) +
    geom_ribbon(data = bs.sint,
                mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = braya_ylabel, x = "Year CE")
brayaInt.plt
```
in the same way

Figure 8 in the manuscript was prepared using
```{r figure-8, fig.show = "hide"}
plot_grid(smallInt.plt, brayaInt.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

```{r profile-range-parameter-in-gp-smooth-model}
nn <- 200    # number of points at which to evaluate profile likelihood
dseq <- seq(15, 500, length.out = nn)  # effective ranges to fit at
Mat <- SEx <- numeric(length = nn)     # object to hold model fits

## loop over the sequence of deparation distances and fit the models
for (i in seq_along(dseq)) { 
    ## iterate over dseq, fit GP GAM w Matérn covariance
    Mat[i] <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, dseq[i])),
                  weights = sampleInterval / mean(sampleInterval),
                  data = braya, method = "REML",
                  family = gaussian())[["gcv.ubre"]]
    ## fit squared exponential
    SEx[i] <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, dseq[i], 1)),
                  weights = sampleInterval / mean(sampleInterval),
                  data = braya, method = "REML",
                  family = gaussian())[["gcv.ubre"]]
}

@@

```{r fit-example-gp-smooths, dependsons = "profile-range-parameter-in-gp-smooth-model"}
## effective range minima from profile likelihood
effRange2 <- with(subset(reml.scr, cor == "Matérn"), dseq[which.min(reml)])
effRange3 <- with(subset(reml.scr, cor == "Exponential"), dseq[which.min(reml)])

## Refit these models: Matern
gp2 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, effRange2)),
           data = braya,
           method = "REML", weights = sampleInterval / mean(sampleInterval))
## Refit these models: Power exponential
gp3 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, effRange3, 1)),
           data = braya,
           method = "REML", weights = sampleInterval / mean(sampleInterval))

## data to predict at
newd <- with(braya, data.frame(Year = seq(min(Year), max(Year),
                               length.out = 1000)))
## create predictions on response scale for both covariance functions
p.gp2 <- transform(newd,
                   fitted = predict(gp2, newdata = newd, type = "response"),
                   effRange = round(effRange2))
p.gp3 <- transform(newd,
                   fitted = predict(gp3, newdata = newd, type = "response"),
                   effRange = round(effRange3))
## joint the two sets of predictions together
pred <- rbind(p.gp2, p.gp3)
## add some categorical variables for plotting
pred <- transform(pred,
                  effRange = factor(effRange),
                  cor = rep(c("Matérn", "Exponential"), each = nrow(newd)))
```

The estimated Gaussian process trends are plotted using

```{r plot-both-gp-smooth-trends}
gp.plt2 <- ggplot(pred, aes(x = Year, y = fitted, colour = cor)) +
    geom_line() + theme(legend.position = "right") +
    geom_point(aes(x = Year, y = UK37), data = braya, inherit.aes = FALSE) +
    scale_colour_manual(name = "", values = c("#e66101","#5e3c99")) +
    labs(y = braya_ylabel, x = "Year CE")
gp.plt2
```

@@

```{r figure-12, fig.show = "hide"}
plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"),
          align = "hv", axis = "lr")
```

@@

```{r adaptive-smooth}
## Adaptive spline, weights as sampleInterval
mod_ad <- gam(UK37 ~ s(Year, k = 45, bs = "ad"), data = braya,
              method = "REML",
              weights = sampleInterval / mean(sampleInterval))
```

# Comparing trends

@@

1. Thin plate regression spline (TPRS),
2. Gaussian process spline (Matérn correlation functions), and
3. Adaptive smoother

@@

```{r braya-so-model-comparisons, dependson = "fit-example-gp-smooths"}
## effective range parameter from profile-likelihood of Matern model
effRange <- effRange2

## TPRS, weights as sampleInterval
mod_tprs <- gam(UK37 ~ s(Year, k = 45, bs = "tp"), data = braya,
                method = "REML",
                weights = sampleInterval / mean(sampleInterval))

@@

```{r, dependson = -1}
## wrapper to simulate from a fitted GAM with the required arguments
simulateSWModels <- function(mod, newdata, nsim, seed = 42) {
    sims <- simulate(mod, nsim = nsim, newdata = newdata, seed = seed)
    as.vector(sims)
}

@@

## align all plots vertically
plots <- align_plots(plt1, plt2, plt3, align = 'v', axis = 'l')

## create the two rows of figures, from `plots`
top_row <- plot_grid(plots[[1]], NULL, ncol = 2, labels = "a")
bot_row <- plot_grid(plots[[2]], plots[[3]], ncol = 1, labels = c("b", "c"))

## combine the two rows, top row has 1 plot row, bottom row has 2, hence
## the rel_heights to even this out
plot_grid(top_row, bot_row, ncol = 1, rel_heights = c(0.5, 1))
```
